\chapter{Conclusion}\label{sec:conc}

We now turn our attention to a more coalesced view of the three type disciplines considered in this article. We first compare different aspects of the four presented type systems, after which we state some concluding remarks. Finally, we present some considerations about steps we believe would be good candidates for future work, based on shortcomings of the presented type systems.

\input{sections/conclusion/comparison}

\section{Concluding remarks}
We have presented and discussed several relevant type disciplines and type systems from the past twenty years of research into type-based static parallel complexity analysis of message-passing processes. Specifically, we have considered type systems based on usages, a Curry-Howard interpretation of temporal session types and sized types to bound the parallel complexity. We have proved a type preservation result of the temporal session type system, and conjectured that the type system provides an upper bound on the span of a well-typed process. We have informally compared the expressiveness and precision of bounds of these type systems, and discussed the prospects of introducing time reconstruction for the corresponding type disciplines. An immediate difficulty in comparing research within this domain, is the variety of message-passing process languages for which these type systems are defined. Therefore, we have introduced a unified $\pi$-calculus language for which only smaller modifications were required amongst the different type systems. We have found that usages are very flexible with respect to expressiveness, as both strong and weak bounds on communication delays can be specified. The Curry-Howard interpretation of temporal session types imposes an interesting constraint on which processes are well-typed. That is, due to composition plus hiding, two processes may only communicate using one name, and so a recursive program must be written as a chain of processes each using a session on one name and providing one on another. With respect to precision, we have found that both usages and session types enable precise bounds on synchronizations on channels. An important distinction is that usages more so describe the capabilities of channels, whereas session types prescribe protocols. We have found sized types to be necessary to obtain precise bounds on recursion. Type inference algorithms for type-based static parallel complexity analysis of message-passing processes have yet to be introduced, however, the type systems we have considered are all built on type disciplines for which such algorithms exist.

\input{sections/conclusion/furtherwork}